---
title: "STAT/MATH 495: Problem Set 09"
author: "Harrison Marick"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

library(tidyverse)
```



# Collaboration

Please indicate who you collaborated with on this assignment: 

I ride solo.


# Question 1: Run k-means

```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.
1. Run KMC 10 times on `observations_2` and comment on the consistentcy of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.

**Answers**:

1. Very inconsistent. Cluster centers seem to be very different with each iteration.
1. The points are scattered throughout, and there is no "natural" dividing point.
1. Very consistent. Little change from trial to trial.
1. Looking at the scatterplot, we can see two distinct groups of points.



# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r}
library(proxy)

c1=runif(2, 0, 2) #randomly select first center (x1, x2 ~ U(0,2))
c2=runif(2, 0, 2) #randomly select second center
centers=data_frame(x1=c(c1[1], c2[1]), x2=c(c1[2], c2[2]))
dist1=sqrt(8) #max possible distance
dist2=sqrt(8) #max possible distance)
while ((dist1+dist2)>0.01){ #threshold of movement between 
points=observations_2
distance_matrix <- proxy::dist(x=points, y=centers)
#distance_matrix
points$cluster=as.factor(apply(distance_matrix, 1, which.min))
c1=c(mean(filter(points, cluster==1)$x1), mean(filter(points, cluster==1)$x2)) #new center1
c2=c(mean(filter(points, cluster==2)$x1), mean(filter(points, cluster==2)$x2)) #new center2

centers_new=data_frame(x1=c(c1[1], c2[1]), x2=c(c1[2], c2[2])) #store new center
dist1<-sqrt((centers$x1[1]-centers_new$x1[1])^2+(centers$x2[1]-centers_new$x2[1])^2) 
#how far did first cluster center move?
dist2<-sqrt((centers$x1[2]-centers_new$x1[2])^2+(centers$x2[2]-centers_new$x2[2])^2)
#how far did second cluster center move?
centers=centers_new
}
centers$cluster=as.factor(c(1,2))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=points) +
  geom_point(data=centers, size=5) + ggtitle("From Scratch Clustering")

```
